---
title: "Mini-Lab - Data Quality"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```


# Introduction

Now is a chance to practice defining, documenting and coding our own data quality rules.

Have a look at the `contractsregister.csv` file which comes from [Aberdeen Council's open data portal](https://www.aberdeenshire.gov.uk/online/open-data/).

# MVP - Defining and coding data quality rules
**Task - 1 hour**

## Business Statement

We are going to start by writing 3 business statements. Ensure they explain the required rule and the context as to why this is important for the business.

1. Write a statement ensuring the DateOfAward is populated

2. Write a statement ensuring that DateOfAward is in the past and is formatted correctly

3. Write a statement ensuring that the ContractValue is a positive numeric value

## Rule Specification

Now for each business statement write rule(s) explaining the test that is being carried out. Ensure you can map each rule to the DQ dimension it is checking.

<details>
<summary>**Hint**</summary>

You may need more than one rule per business statement

</details>


## The coded rules

For each data quality rule defined above, try and write `R-code` that can be used to test that rule.

You may want to use this function, which checks if a date is invalid:

```{r}
is_invalid_date <- function(date){
 
  parsed_date <- try(as.Date(date, format= "%d/%m/%Y"))
  
  return(inherits(parsed_date, "try-error") | is.na(parsed_date))
}
```

And this function, which checks if a date is in the past:

```{r}
is_in_past <- function(date){
  
  return(lubridate::dmy(date) <= lubridate::today())
  
}
```

Don't worry if you don't understand the code! We'll learn about dates and times in week 3.

## Summary statistics 

Create summary statistics, showing the volumes that pass and fail each of the rules you have defined.








