---
title: "Python Machine Learning - Homework"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r, echo = FALSE}
library(reticulate)
use_python("/anaconda3/bin/python")

knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

# MVP

Use the forest fire data from last night's homework.

We want you to build a linear regression model to predict the size of a forest fire in both in *Python* and in *R*. No need to explore the data first, since you did that yesterday!

1. Create a model in Python
2. Create a model in R.
3. Write a short paragraph comparing what was easy and what was difficult in both languages.
4. Interpret the model.

```{python}
# Python version

import pandas as pd
from sklearn.linear_model import LinearRegression
from sklearn.model_selection import train_test_split


forest_fire = pd.read_csv("forestfires.csv")

forest_fire = forest_fire.dropna()

forest_fire = pd.get_dummies(forest_fire, drop_first = True)

y = forest_fire["area"]
X = forest_fire.drop(columns = "area")

X_train, X_test, y_train, y_test = train_test_split(X, y, test_size = 0.1, random_state = 7)

model = LinearRegression()

model.fit(X_train, y_train)

model.score(X_test, y_test)
```

```{r}
# R version

library(readr)

forest_fire <- read_csv("forestfires.csv")

model <- lm(area ~ ., data = forest_fire)

summary(model)
```


# Extension

Containing in whatever language you like, try to improve this model. 


```{r}
# Predicing log of area is much more effective. Lots of other improvements are possible! 

model <- lm(log(area + 1) ~ ., data = forest_fire) # Need log + 1 because of 0 values

summary(model)
```

