---
title: "`Dplyr` Homework with answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>
The data for this homework comes from an Open Data Science website called [Kaggle](https://www.kaggle.com/). Kaggle has many open source datasets for you to use and most of them come with code uploaded by other users showing different ways to explore the data. It is a great way to learn about data-wrangling and analysis and if you are interested then set up your own account and get going. 

For this assignment we will make use of their **Starbucks Calorie** dataset. You can find out more information about each dataset and what each column represents **[here](https://www.kaggle.com/starbucks/starbucks-menu)**, **but we have uploaded the version you need to your classnotes already**.   


```{r, warning=FALSE, message=FALSE, echo = FALSE}
library(tidyverse)
library(nycflights13)
```

<br>

# MVP

<br>

**Question 1**

Load in the dataset and call it `drinks_content`, eyeball it by printing it out (first 5 rows, last 5 rows, and then all of the data) and find out the dimensions (number of rows/columns) and variable names. 
<br>

<details>
<summary> **Answer** </summary>
```{r}
# load the data in 
drinks_content <- read_csv("data/starbucks_drinkMenu_expanded.csv")

# first 5 rows
head(drinks_content,5)

# last 5 rows
tail(drinks_content,5)

# print the data out 
drinks_content

# get the dimensions
dim(drinks_content)

# get variable names 
names(drinks_content)
```
</details>
<br>


**Question 2**

Calculate some quick summary statistics to check the mean, and check the normality of the `Calories` variable. Write a quick summary of your initial impressions of the data. 
<br>
<details>
<summary> **Answer** </summary>
```{r}

# check the descriptive stats
summary(drinks_content$Calories)

# histogram
hist(drinks_content$Calories)
```
</details>
<br>


**Question 3**

Check if you have any outliers in the Calories variable by creating a boxplot. Then get the actual values of any outliers you have. Finally, decide what you want to do here. Do you want to remove the outliers? Do you want to keep them? Justify your decision either way. As you work through your outlier identification and removal process, write a summary of what you find (i.e. do you find any outliers), what you decide to do (why you are removing them or not, what criteria you might use), and how you removed them if you did (e.g. did you remove all or not). 
<br>

<details>
<summary> **Answer** </summary>

```{r}
# make a boxplot of outliers
boxplot(drinks_content$Calories)
## yep, seems I have an outlier... 

# get the outliers 
outliers <- boxplot(drinks_content$Calories, plot=FALSE)$out
print(outliers)
## I only have one, and it's pretty high, BUT... some drinks probably have loads more calories than others. So i'll keep it in. 

# # if you did remove it, you might do it this way:
# # remove them 
# drinks_content[which(drinks_content$Calories %in% outliers),]
# 
# # Now you can remove the rows containing the outliers, one possible option is:
# drinks_content_nooutlier <- drinks_content[-which(drinks_content$Calories %in% outliers),]
# 
# # check now outliers are gone
# boxplot(drinks_content_nooutlier$Calories)
```
</details>
<br>


**Question 4**

Select the variables `beverage_category`, `beverage`, `beverage prep` and `calories` from your data. Since we are interested in the calorie content, check if there are any NA values in the data, and drop them if there are. 
<br>

<details>
<summary> **Answer** </summary>
```{r}
# select the variables
drinks <- select(drinks_content, Beverage_category, Beverage, Beverage_prep, Calories)

# check for NA's
is.na(drinks$Calories)

# no NA values, so we're all good
```
</details>
<br>


**Question 5**

Filter the data so we only take "classic espresso drinks".
<br>

<details>
<summary> **Answer** </summary>
```{r}

# filter the data 
espresso_drinks <- filter(drinks, Beverage_category == "Classic Espresso Drinks")

```
</details>
<br>


**Question 6**

Create a new tibble called `drink_calories` which contains a variable called `calorie_diff`. `calorie_diff` should store the difference between 135 calories (135 calories = 10 cubes of sugar!) and the calories in each drink. 
(hint: you'll want to subtract the calories from 135 to see which drink have more or less than 10 cups of sugar). 
<br>

<details>
<summary> **Answer** </summary>
```{r}
# create a new variable 
drink_calories <- mutate(espresso_drinks, calorie_diff = 135 - Calories)
```
</details>
<br>


**Question 7**

Create a new tibble called `drinks_grouped` where you group your data by the type of beverage prep. Then summarise the mean calories in each drink (by group) in a variable called `mean_cal`, and save it to `drinks_summary`. Finally arrange this summary data in descending order of mean calories. 
<br>

<details>
<summary> **Answer** </summary>

```{r}
# group the data 
drinks_grouped <- group_by(drink_calories, Beverage_prep)

# summarise the data 
drinks_summary <- summarise(drinks_grouped, mean_cal = mean(Calories))

# arrange in descending order 
arrange(drinks_summary,desc(mean_cal))

```
</details>
<br>


**Question 8**

Hopefully you can see by now how much hassle it can be to constantly keep creating tibbles to store all your data. 
Finish up by using the `%>%` operator to put all the answers you've got from questions 4-7 into one pipe. You don't need to include the check for NA values in this, as you'd typically check that before you start doing any analysis. 
<br>

<details>
<summary> **Answer** </summary>
```{r}

# pipe everything 
drinks_pipe <- drinks_content %>%
    select(Beverage_category, Beverage, Beverage_prep, Calories) %>%
    filter(Beverage_category == "Classic Espresso Drinks") %>%
    mutate(calorie_diff = 135 - Calories) %>%
    group_by(Beverage_prep) %>%
    summarise(mean_cal = mean(Calories)) %>%
    arrange(desc(mean_cal))

```
</details>
<br>


**Question 9**

Use the `drink_calories` data to find the two drinks (`Beverage`) with least and most Calories. *(Hint: use the `min_rank()` function)*       
<br>

<details>
<summary> **Answer** </summary>
```{r}

# calculate least calories
least_calories <- drink_calories %>%
  arrange(Calories) %>%
  mutate(ranking = min_rank(Calories)) %>%
  filter(ranking <=2)

# calculate most calories
most_calories <- drink_calories %>%
  mutate(ranking = min_rank(desc(Calories))) %>%
  filter(ranking <=2)

```
</details>
<br>


**Question 10**

For each drink in the `drinks_grouped` data, find all drinks which more calories than the average drink. Save this in a new tibble called `above_average_calories`. 
<br>

<details>
<summary> **Answer** </summary>
```{r}
above_average_calories <- drinks_grouped %>%
  filter(Calories > mean(Calories))
```
</details>
<br>


**Question 11**

Using the `drinks_grouped` data, apply the `cume_dist` function to get the top 25% of Calorie drinks within each group. Save this in a new tibble called `drinks_distribution`.
<br>

<details>
<summary> **Answer** </summary>
```{r}
drinks_distribution <- drinks_grouped %>%
  filter(cume_dist(desc(Calories)) < 0.25) %>%
  select(Beverage_category,Beverage_prep)

drinks_distribution

names(drinks_distribution)
```
</details>
<br>


# Extension : practice your joins
<br>

For the extension task, we will use the `nycflights13` package data to work on some joins. This dataset contains all 336776 flights that departed from New York City in 2013. The data comes from the US Bureau of Transportation Statistics, and is documented in `?nycflights13`. It contains 5 related tables: 
<br>
* airlines : airline names
* airports : airport metadata
* flights  : flight time data for NYC departing flights 
* planes   : plane metadata
* weather  : hourly weather data 

[You can find out more about the package here]( https://CRAN.R-project.org/package=nycflights13)  
<br>

You might need to install the package first. 

<br>

**Extension question 1**

Add the weather to the flights data by city of origin.  
<br>

<details>
<summary> **Answer** </summary>
```{r}
weather_join <- left_join(flights, weather, by=c("origin","time_hour"))

```
</details>
<br>


**Extension question 2**

Add the location of the origin and destination (i.e. the `lat` and `lon`) to flights. Only keep those that have a location. 
<br>
<details>
<summary> **Answer** </summary>
```{r}
location <- flights %>%
     inner_join(airports, by=c("origin"="faa")) %>%
     inner_join(airports, by=c("dest"="faa"))

```
</details>
<br>


**Extension question 3**

Find the 10 most popular destinations to fly to, and then perform a join to find which flight went with those top 10 destinations. 
<br>

<details>
<summary> **Answer** </summary>
```{r}

# find the top 10 destinations
top_destinations <- flights %>%
  group_by(dest) %>%
  summarise(flight_cnt = n()) %>%
  arrange(desc(flight_cnt)) %>%
  filter(row_number() <= 10) 

# could use a semi_join or an inner join to only keep flights destination that match the top 10
flights %>% 
  semi_join(top_destinations)

```
</details>
















