---
title: "Advanced query syntax part 1 - WHERE"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

<div class="blame">
  author: "Del Middlemiss"<br>
  date: "9th June 2019"
</div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```
# Learning Objectives<br>

* Be able to use additional comparison operators, and `AND` and `OR` combinations
* Write queries using `BETWEEN`, `NOT` and `IN`
* Understand and be able to use inexact comparisons with `LIKE` and wildcards
* Understand `IS NULL`

**Duration - 45 minutes**<br>

As stated earlier, data analysts more often read records in databases than create, update or delete them (recall CRUD). So, it makes sense to focus effort on learning `SELECT` syntax beyond simple `WHERE` clauses. We'll work through a series of examples, continuing to use the `omni_employees` local PostgreSQL database we set up in the last lesson. 

We'll come back later in the week to look at how we manipulate the returned data from a query.

<hr>

# Setup

First, we need to re-establish the `R` connection to the database.

```{r}
library(keyring)
library(RPostgreSQL)

keyring_unlock("local")
username <- key_get("postgresql_username", keyring = "local") 
password <- key_get("postgresql_password", keyring = "local")
keyring_lock("local")
db_connect <- dbConnect(drv = PostgreSQL(), user = username, password = password, dbname = "omni_employees", host = "localhost")
rm(username, password)
```

Just so we don't forget, add in the disconnect statement at the end of the file, and then put our code in between

```{r, eval = FALSE, echo = TRUE}
# keep moving this down, put code in between
dbDisconnect(db_connect)
```

<hr>

# Simple `WHERE` clauses

So far we've seen pretty simple `WHERE` clauses, e.g. find the employee with `id` equal to 3.

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE id = 3")
```

<hr>

# Additional comparison operators

OK, but what about this problem:<br>

<div class='emphasis'>
"Find all the employees working 0.5 full-time equivalent hours or greater."
</div>

<br>
We can solve this using operators other than `=`
<br>

| operator | meaning |
| --- | --- |
| != | not equal to |
| > | greater than |
| < | less than |
| >= | greater than or equal to |
| <= | less than or equal to |

<br>
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE fte_hours >= 0.5")
```

<blockquote class='task'>
**Task - 2 mins** Write and execute a query answering this problem:
<center>"Find all the employees not based in Brazil."</center>
<details>
<summary>**Solution**</summary>
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country != 'Brazil'")
```
</details>
</blockquote>

<hr>

# `AND` and `OR` 

If required, we can create more complex clauses using the `AND` and `OR` operators

<div class='emphasis'>
"Find all employees in China who started working for OmniCorp this year".
</div><br>

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country = 'China' AND start_date >= '2019-01-01'")
```

Sometimes we have to be careful with the **order of evaluation** of conditions. Consider the following example

<br>
<div class='emphasis'>
"Find all employees in China who either started working for OmniCorp this year or who are enrolled in the pension scheme".
</div><br>

The logic of this as written is fairly clear. All the returned employees should be based in China **and** they should either have started work for OmniCorp this year **or** be enrolled in the pension scheme. Let's have a first go at writing this!

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country = 'China' AND start_date >= '2019-01-01' OR pension_enrol = TRUE")
```

Argh! We get employees from countries other than China! What's gone wrong? We need to worry about the order of evaluation. We want the `OR` operation to execute **before** the `AND` operation, and we enforce this by use of parentheses!

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country = 'China' AND (start_date >= '2019-01-01' OR pension_enrol = TRUE)")
```

That's better! Let's see a few more examples using `AND` and `OR`

<br>
<div class='emphasis'>
"Find all employees who work between 0.25 and 0.5 full-time equivalent hours inclusive".
</div><br>

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE fte_hours >= 0.25 AND fte_hours <= 0.5")
```

<br>
<div class='emphasis'>
"Find all employees who started working for OmniCorp in years other than 2017".
</div><br>

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE start_date < '2017-01-01' OR start_date > '2017-12-31'")
```

<hr>

# `BETWEEN`, `NOT` and `IN`

The syntax in the last two examples is clumsy: it would be better to be able to define a range and then `SELECT` for records with fields in/not in that range. The `BETWEEN` keyword lets us do just that! 

Let's rewrite the previous two queries using `BETWEEN`:

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE fte_hours BETWEEN 0.25 AND 0.5")
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE start_date NOT BETWEEN '2017-01-01' AND '2017-12-31'")
```

Note these two points:

* the range defined by `BETWEEN` is **inclusive** of the end points. So, in the first example, records with `fte_hours` of exactly 0.2 or 0.5 will be selected
* we use the `NOT BETWEEN` combination in the second example to select all records with `start_date` not in the range (where, again, the end points are included in the range).

<blockquote class='task'>
**Task - 5 mins** See if you can work out the query syntax to answer this problem:
<center>"Find all employees who started work at OmniCorp in 2016<br> who work 0.5 full time equivalent hours or greater."</center>

<details>
<summary>**Hint**</summary>
You need an `AND` combination of two conditions, one of which involves `BETWEEN`
</details>
<details>
<summary>**Solution**</summary>
```{r}
# parentheses might help you to see different logical 'levels' of the conditions
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE (start_date BETWEEN '2016-01-01' AND '2016-12-31') AND (fte_hours >= 0.5)")
```
</details>
</blockquote>
<hr>

The `IN` operator helps us deal with discrete valued fields more efficiently. Consider:

<div class='emphasis'>
"Find all employees based in Spain, South Africa, Ireland or Germany."
</div>

The long-winded way to write this is
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country = 'Spain' OR country = 'South Africa' OR country = 'Ireland' OR country = 'Germany'")
```

or using `IN`
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country IN ('Spain', 'South Africa', 'Ireland', 'Germany')")
```

We can also use `NOT` with `IN`

<div class='emphasis'>
"Find all employees based in countries other than Finland, Argentina or Canada."
</div>

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country NOT IN ('Finland', 'Argentina', 'Canada')")
```

<hr>

# `LIKE` and wildcards

Your manager comes to you and says <br>

<div class='emphasis'>
"I was talking with a colleague from Greece last month, I can't remember their last name exactly, I think it began 'Mc...' something. Can you find them?"
</div><br>

You can see we're dealing with an **inexact comparison** here. How do we do this? Using the `LIKE` operator with a **wildcard**.

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE country = 'Greece' AND last_name LIKE 'Mc%'")
```
<br>

Here are the wildcards we can use: <br>

| wildcard | meaning |
| --- | --- |
| _ | a single character |
| % | a collection of characters |

<br>

We can place wildcards **anywhere** inside the string in the condition:<br>

<div class='emphasis'>
"Find all employees with last names containing the phrase 'ere' anywhere"
</div>

<br>
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE last_name LIKE '%ere%'")
```

<br>
<div class='emphasis'>
"Find all employees in the Legal department with a last name beginning with 'D'" 
</div>

<br>
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE department = 'Legal' AND last_name LIKE 'D%'")
```

<blockquote class='task'>
**Task - 2 mins** Write a query using `LIKE` and wildcards to answer:
<center>"Find all employees having 'a' as the second letter of their first names."</center> 

<details>
<summary>**Hint**</summary>
You can use a '_' wildcard for the first letter of `first_name`.
</details>

<details>
<summary>**Solution**</summary>
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE first_name LIKE '_a%'")
```
</details>
</blockquote>

`LIKE` distinguishes between capital and lower case letters. If we need a case-insensitive version, we can use `ILIKE`.

<hr>

# `IS NULL`

<br>
<div class='emphasis'>
"We need to ensure our employee records are up-to-date. Find all the employees who do not have a listed email address."
</div>

<br>
We can use the `IS NULL` condition to fulfill this request.

```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM employees 
           WHERE email IS NULL")
```

We can also use the `NOT` operator here too: `IS NOT NULL` is a valid condition!

<hr>

# Recap - Full query syntax

In summary, here are the different keyword components of a `SELECT` query, the order in which they must appear, and whether they are required or optional <br>

This table shows where we've got to:<br>

| Order | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 1 | SELECT | Column to query | Yes |
| 2 | FROM | Table to query |  Yes |
| 3 | WHERE | Row-level filter | No |

<br>
while here's a look ahead at what's still to come:<br>

| Order | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 5 | GROUP BY| Grouping for aggregates | No |
| 6 | HAVING | Group-level filter | No |
| 7 | ORDER BY | Sort order | No |
| 8 | LIMIT | How many records to return | No |






