---
title: "Task 1 - Analysis results "
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Loading in data 

```{r}
library(tidyverse)
library(here) 

# Use the 'here' function so that path directories work in notebooks the same way the work in scripts
decathalon <- read_csv(here("clean_data/decathlon_clean.csv"))
```

#  Question 1
Who had the longest long jump seen in the data?

```{r}
decathalon %>%
  filter(event == "long_jump") %>%
  arrange(desc(score)) %>%
  head(1)
```


# Question 2
What was the average 100m time in each competition?

```{r}
decathalon %>%
  filter(event == "running_100m") %>%
  group_by(competition) %>%
  summarise(av_time = mean(score))
```


# Question 3 
Who had the highest total points across both competitions?

```{r}
decathalon %>%
  distinct(surname, competition, points) %>%
  group_by(surname) %>%
  summarise(points_total = sum(points)) %>%
  arrange(desc(points_total)) %>%
  head(1)
```


# Question 4
What was the shot-put scores for the top three competitors in each competition?

```{r}
decathalon %>%
  filter(rank <= 3) %>%
  filter(event == "shot_put")
```


# Question 5
What was the average points for competitors who ran the 400m in less than 50 seconds vs. those than ran 400m in more than 50 seconds?

```{r}
decathalon %>%
  filter(event == "running_400m") %>%
  mutate(fast_400 = score < 50) %>%
  group_by(fast_400) %>%
  summarise(avg_points = mean(points))
```


