---
title: "Reading Data Lab answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../resources/note_styles.css
  pdf_document: default
---

<div class="blame">
author: "Mhairi McNeill"<br>
date: "08/07/2019"
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE)
```

# Flat Files

Read in the following datasets. You may need define the variable types, skip rows or give your own variable names in each of them.

Before you read in each file, either view it in the Terminal (`less`) or open it in a text editor, so you can see what type it is. 

1. `user-for-x.txt`
2. `female_names.txt`
3. `edinburgh-trees.csv`

```{r}
library(readr)
uber_for_x <- read_tsv("data/uber-for-x.txt")
```
```{r}
uber_for_x
```


```{r}
female_names <- read_lines("data/female_names.txt")
```
```{r}
female_names
```


```{r}
edinburgh_trees <- read_csv("data/edinburgh-trees.csv", skip = 9, col_names = c("Latin Name", "Common Name", "Location", "Lat/Lon", "Age"))
```
```{r}
edinburgh_trees
```


```{r}
# You could read the names as a separate data-frame and use that to name the full data. Something like this:
tree_names <- read_csv("data/edinburgh-trees.csv", skip = 2, n_max = 5, col_names = c("letter", "name"))

trees <- read_csv("data/edinburgh-trees.csv", skip = 9, col_names = tree_names$name)
```
```{r}
trees
```

# Excel Files

Can you figure out how to read all the sheets from `edinburgh_seedmix.xlsx` into one data frame?

<details>
<summary>**Hint**</summary>

Look back at your notes on loops to see an example of reading multiple files in a loop.

</details>

```{r}
library(readxl)

seedmix <- NULL

n_sheets <- length(excel_sheets("data/edinburgh_seedmix.xlsx"))

for (sheet in 1:n_sheets) {
  seedmix_part <- read_excel("data/edinburgh_seedmix.xlsx",
    sheet = sheet,
    col_names = c(
      "park",
      "location",
      "m2",
      "grams",
      "seed_mix",
      "species"
    )
  )
  seedmix <- rbind(seedmix, seedmix_part)
}

seedmix
```


Take the `populism` dataset that we read in earlier and write it to an Excel file. In the Excel file, you want the data for each region in a different sheet.

<details>
<summary>**Hint 1**</summary>
Use a loop to go through each part of the data frame.
</details>
<details>

<summary>**Hint 2**</summary>
Use `subset()` to extract only a part of the data frame.
</details>

```{r}
library(openxlsx)
populism <- read_excel('data/populism.xlsx')

# create your workbook
wb <- createWorkbook()

for (reg in unique(populism$region)) {
  region_sheet <- subset(populism, region == reg)
  
  # add a sheet
  addWorksheet(wb, sheetName = reg)
  writeData(wb, sheet = reg, x = region_sheet)
}

saveWorkbook(wb, file = "populism_loop.xlsx")

```

# Other File Formats

1. Read in `golf.sas7bdat`.

2. What kind of file is this?

3. Convert all the labelled values to their underlying numeric value.

4. Convert all the labelled values to factors.

5. What do you notice about this conversion?

```{r}
library(haven)

golf <- read_sas("data/golf.sas7bdat")
zap_labels(golf)
# This is a file from SAS
```
```{r}
as_factor(golf)

# Just the same as golf! This dataset doesn't use labelled values, even though they are avalible in SAS.
```

# JSON

Parse the file `starwars.json` as:

1. a `list`
2. a `data.frame`


```{r}
library(jsonlite)

raw_starwars <- read_file("data/starwars.json")
starwars <- parse_json(raw_starwars)
```


```{r}
starwars_df <- fromJSON(raw_starwars)
```



