---
title: "Data Types Homework - Answers"
output:
  html_document: 
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Make sure you create a new project in which to do this homework!

# Part 1

In this part, you will compare several ways of storing the same data. 

<br>
**Question 1.**  

Create a vector called `stock_prices` with the following data points: 23, 27, 23, 21, 34, 24, 25, 30, 37, 30.

Create a vector called `day_of_week`, which repeats Monday to Friday twice. i.e.
'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'.

```{r}
stock_prices <- c(23, 27, 23, 21, 34, 24, 25, 30, 37, 30)
day_of_week <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
```

<br> 

**Question 2.**

Create a vector called `day_of_week_factor`, which has the same data as `day_of_week`, but is of `factor` type.

Created a vector called `named_stock_prices`, which contains the stock prices, named by days of the week. 

```{r}
day_of_week_factor <- as.factor(day_of_week)
```

```{r}
named_stock_prices <- stock_prices
names(named_stock_prices) <- day_of_week
```
<br>


**Question 3.**  

Make a data frame called `stock_data` containing the stock price and day of the week information, and `View()` the result.

```{r}
stock_data <- data.frame(stock_prices = stock_prices, day_of_week = day_of_week)
```

```{r, eval=FALSE}
View(stock_data)
```
<br>

**Question 4.**  

Make a list called `stock_list` with named elements as follows: the vector of `stock_prices`, the vector `day_of_week`, and the data frame you made in question 3. Choose sensible names for the list elements. 

```{r}
stock_list <- list(stock_prices = stock_prices, day_of_week = day_of_week, stock_data = stock_data)

```
<br>


# Part 2

In this part, you will practice selecting elements from vectors, data frames, matrices and lists.

For Part 2, you need the code you've written for Part 1. If you've really struggled with this and can't do it, then the code for Part 1 is listed underneath the hidden button below. Do try your hardest first though before using the code below! 

<details>
```{r}
stock_prices <- c(23, 27, 23, 21, 34, 24, 25, 30, 37, 30)

day_of_week <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')

day_of_week_factor <- as.factor(day_of_week)

stock_data <- data.frame(stock_prices = stock_prices, day_of_week = day_of_week)

stock_list <- list(stock_prices = stock_prices, day_of_week = day_of_week, stock_data = stock_data)
```
</details>
<br>

**Question 1.**

Select the 4th, 5th and 6th stock prices from `stock_prices`.

Select the stock prices greater than 30 from `stock_prices`.

```{r}
stock_prices[4:6]
```

```{r}
stock_prices[stock_prices > 30]
```
<br>

**Question 2.**

Select the `day_of_week` column from the `stock_data` data frame.

Select the rows of the `stock_data` data frame where the `stock_prices` are less than 25.

Select the 1st column of `stock_data`.

Select the 3rd row of `stock_data`.

Select the 3rd element in the 1st column of `stock_data`.

```{r}
stock_data$day_of_week
```

```{r}
subset(stock_data, stock_prices < 25)
```

```{r}
stock_data[, 1]
```

```{r}
stock_data[3, ]
```

```{r}
stock_data[3, 1]
```
<br>

**Question 4.** 

Select the data frame from within the list of stocks.

```{r}
stock_list$stock_data
```
<br>


# Extension

Download the `tidyquant` library.

1. Use it to get stock prices from a company of your choice from the current year. You'll need to look up the help file to do this.

2. Edit the data frame to keep only the `date` and `close` columns.

3. Subset to find the rows at which the `close` price is higher than average.


```{r, message=FALSE, warning=FALSE}
library(tidyquant)

google_prices  <- tq_get("GOOGL", get = "stock.prices", from = "2019-01-01")

google_prices[ , c("date", "close")]
  
subset(google_prices, close > mean(close))
```





