---
title: "`Spark` and `sparklyr` setup"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---


# Installing `Java 8`, `sparklyr` and `Spark 2.4.5`

The `sparklyr` package makes installing `Spark` and `Scala` really easy, but first we need to make sure that `Java 8` is installed on your machine.

To be honest, `Java` versioning is a bit of a mess, and if you install multiple `Java` versions on your computer it can be difficult to select the version to use. So we recommend removing `Java 11` if you find it on your system.

* **Check `Java` version**. In the `RStudio` console, run the command `system("java -version")`. 
  - If you see nothing, an error, or a prompt to 'install a JDK', that's fine as it means no `Java` is currently installed - skip to the **Install `Java 8`** point. 
  - If you see a `Java` version that starts with '1.8', that's fine too, as it means `Java 8` is installed - skip to the **Install `sparklyr`** point. 
  - If you see a `Java` version higher than '1.8' you need to uninstall the later version of `Java` - skip to the **Removing later `Java` versions** point.


* **Removing later `Java` versions**. In the `Terminal` `cd /Library/Java/JavaVirtualMachines` and have a look at the directories there (`ls`). These are the various versions of `Java` installed on your machine. You should remove any directories for `Java` versions after '1.8'. For example, if I found a directory `adoptopenjdk-11.jdk`, I would remove it with the command `sudo rm -rf adoptopenjdk-11.jdk` (be careful typing this command). 


* **Install `Java 8`** In the `Terminal`, change to your home directory (`cd ~`) and then execute `brew cask install adoptopenjdk/openjdk/adoptopenjdk8` (it will update Homebrew and may ask for your password). Afterwards, back in the `RStudio` console, try `system("java -version")` again: hopefully you will now see a version starting '1.8'. Try restarting your `R` session and trying again if you see any other output. 


* **Install `sparklyr`** Install the `R` `sparklyr` package in the normal way [e.g. in the `RStudio` console, run `install.packages("sparklyr")`].


* **Use `sparklyr` to install `Spark`** Next we'll use `sparklyr` to install the current stable version of `Spark` (2.4.5). So, in a code block, execute

```{r, eval=FALSE}
library(sparklyr)
spark_install(version = "2.4.5") # this is quite a large download
```

Once that's done, test your `Spark` installation with this code block

```{r, eval=FALSE}
sc <- spark_connect(master = "local")
spark_cars <- copy_to(sc, mtcars)
```

If these run fine you're good to go! **Any problems, we can help you fix them, don't worry!**

