---
title: "SQL lab - solutions"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

<div class="blame">
author: "Del Middlemiss"<br>
date: "16th June 2019"
</div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

**Duration - 120 minutes**

In this lab we'll return again to the `omni_employees` database, but now looking at all tables and the relationships between them.

The `psql` command line interface offers a number of commands to get further information on databases and the tables within them. All `psql` commands start with a backslash `\`.
<br>

| command | output |
| --- | --- |
| `\l` | list databases|
| `\c db_name` | connect to database *db_name* |
| `\dt` | list tables in a database |
| `\d+ table_name` | get detailed information on table *table_name* |

<br>
```bash
# connect to local server
psql 
# list all databases
\l
# connect to omni_employees
\c omni_employees
# list all tables in omni_employees
\dt
# get detailed information on the committees table
\d+ committees
```
<br>

<hr>

# Preparation

* In `psql`, connect to `omni_employees`, list all the tables, and get detailed information on each of the tables using the `\d+` command. 
  - primary keys are shown in the 'Indexes' section.
  - foreign keys are shown in the 'Foreign-key constraints' section.
  - the 'Referenced by' section indicates if the table is the *target* of a foreign key in another table.

* Investigate table relationships: look for primary keys and foreign keys, see if you can spot any one-to-one, one-to-many or many-to-many relationships.

Solution (PK is 'primary key', FK is 'foreign key'):

* `committees`: PK `id`, no FKs, referenced by `employees_committees`.
* `employees`: PK `id`, FKs `pay_detail_id` and `team_id`, referenced by `employees_committees`.
* `employees_committees`: PK `id`, FKs `committee_id` and `employee_id`, not reference by any other table.
* `pay_details`: PK `id`, no FKs, referenced by `employees`.
* `teams`: PK `id`, referenced by `employees`.
<br>
Relationships: 

* one employee has one pay_detail
* one team has many employees, one employee has one team
* one committee has many employees, one employee has many committees

# MVP

Setup a connection from R to your local `omni_employees` database, then write **a single query** to solve each of the following problems:

```{r}
library(keyring)
library(RPostgreSQL)

keyring_unlock("local")
username <- key_get("postgresql_username", keyring = "local") 
password <- key_get("postgresql_password", keyring = "local")
keyring_lock("local")
db_connect <- dbConnect(drv = PostgreSQL(), user = username, password = password, dbname = "omni_employees", host = "localhost")
```

<br>

1. Find all the `employees` who work in the 'Human Resources' department.
```{r}
dbGetQuery(db_connect, 
          "SELECT * 
          FROM employees
          WHERE department = 'Human Resources'")
```

<br>

2. Get the first name, last name, and country of the `employees` who work in the 'Legal' department
```{r}
dbGetQuery(db_connect, 
          "SELECT first_name, last_name, country 
          FROM employees
          WHERE department = 'Legal'")
```

<br>

3. Count the number of `employees` based in Portugal.
```{r}
dbGetQuery(db_connect, 
          "SELECT COUNT(id)
          FROM employees
          WHERE country = 'Portugal'")
```

<br>

4. Count the number of `employees` based in either Portugal or Spain.
```{r}
dbGetQuery(db_connect, 
          "SELECT COUNT(id)
          FROM employees
          WHERE country IN ('Portugal', 'Spain')")
```

<br>

5. Count the number of `pay_details` records lacking a `local_account_no`
```{r}
dbGetQuery(db_connect, 
          "SELECT COUNT(id)
          FROM pay_details
          WHERE local_account_no IS NULL")
```

<br>

6. Are there any `pay_details` record lacking both a `local_account_no` and `iban` number?
```{r}
dbGetQuery(db_connect, 
          "SELECT COUNT(id)
          FROM pay_details
          WHERE local_account_no IS NULL AND iban IS NULL")
# count is zero, so no
```

<br>

7. Get a list of `employees` first names and last names ordered alphabetically by their last name (put any nulls last).
```{r}
dbGetQuery(db_connect,
           "SELECT first_name, last_name
           FROM employees
           ORDER BY last_name ASC NULLS LAST")
```

<br> 

8. Get a list of `employees` first names, last names and countries, ordered alphabetically first by their country and second by their last name (put any nulls last).
```{r}
dbGetQuery(db_connect,
           "SELECT first_name, last_name, country
           FROM employees
           ORDER BY country ASC NULLS LAST, last_name ASC NULLS LAST")
```

<br>

9. Find the top ten highest paid `employees` in the corporation.
```{r}
dbGetQuery(db_connect,
           "SELECT *
           FROM employees
           ORDER BY salary DESC NULLS LAST
           LIMIT 10")
```

<br>

10. Find the first name, last name and salary of the lowest paid of the `employees` in Hungary 
```{r}
dbGetQuery(db_connect,
           "SELECT first_name, last_name, salary
           FROM employees
           WHERE country = 'Hungary'
           ORDER BY salary ASC NULLS LAST
           LIMIT 1")
```

<br>

11. How many `employees` have a first name beginning with 'F'?
```{r}
dbGetQuery(db_connect,
           "SELECT COUNT(id)
           FROM employees
           WHERE first_name ILIKE 'F%'")
```

<br>

12. Find all the details of any `employees` with a 'yahoo' e-mail address?
```{r}
dbGetQuery(db_connect,
           "SELECT *
           FROM employees
           WHERE email ILIKE '%yahoo%'")
```

<br>

13. Count the number of pension enrolled `employees` not based in either France or Germany.
```{r}
dbGetQuery(db_connect,
           "SELECT COUNT(id)
           FROM employees
           WHERE pension_enrol IS TRUE AND country NOT IN ('France', 'Germany')")
```

<br>

14. Breakdown the numbers of `employees` enrolled, not enrolled and whose enrollment status is unknown in the corporation pension scheme.
<details>
<summary>**Hint**</summary>
'Breakdown the numbers' indicates a `GROUP BY` operator is needed.
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT pension_enrol, COUNT(id)
           FROM employees
           GROUP BY pension_enrol")
```

<br>

15. What is the maximum salary among `employees` in the Engineering department who work 1.0 full-time equivalent hours?
```{r}
dbGetQuery(db_connect,
           "SELECT MAX(salary)
           FROM employees
           WHERE department = 'Engineering' AND fte_hours = 1.0")
```

<br>

16. Obtain a count by department of the `employees` who started work with the corporation in 2003.
```{r}
dbGetQuery(db_connect,
           "SELECT department, COUNT(id) AS num_employees
           FROM employees
           WHERE start_date BETWEEN '2003-01-01' AND '2003-12-31'
           GROUP BY department")
```

<hr>

# Extensions

Some of these problems may need you to do some online research on SQL statements we haven't seen in lessons... Don't worry, we'll give you pointers and it's good practice!

<br>

1. Show the average salary in any countries in which more than 30 `employees` are based. Order the list descending by average salary.
<details>
<summary>**Hints**</summary>
    * A `HAVING` clause is needed to filter using an aggregate function.
    * You can pass a column alias to `ORDER BY`
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT country, count(id) AS num_employees, AVG(salary) AS average_salary
           FROM employees
           GROUP BY country
           HAVING COUNT(id) > 30
           ORDER BY average_salary DESC")
```

<br>

2. Find the first name and last name of all `employees` who lack a `local_tax_code`.
<details>
<summary>**Hint**</summary>
`local_tax_code` is a field in table `pay_details`, and `first_name` and `last_name` are fields in table `employees`, so this query requires a `JOIN`!
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT first_name, last_name
           FROM employees AS e INNER JOIN pay_details AS pd
           ON e.pay_detail_id = pd.id
           WHERE pd.local_tax_code IS NULL")
```

<br>

3. Find the first name, last name and team name of `employees` who are members of `teams` for which the charge cost is greater than 80. Order the `employees` alphabetically by last name.
<details>
<summary>**Hint**</summary>
`charge_cost` may be the wrong type to compare with value `80`. Can you find a way to convert it **without changing** the database?<br>
This query will require a `JOIN`, with a `WHERE` clause on a field in `teams`.
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT e.first_name, e.last_name, t.name AS team_name
           FROM employees AS e INNER JOIN teams AS t
           ON e.team_id = t.id
           WHERE CAST(t.charge_cost AS INT) > 80
           ORDER BY e.last_name  ASC NULLS LAST")
```

<br>

4. Breakdown the number of `employees` in each of the `teams`, including any `teams` without members. Order the list by increasing size of team.
<details>
<summary>**Hints**</summary>
    * Two tables involved here, so a `JOIN` is necessary.
    * We want to include **all teams** (not just those with member employees), so an `INNER JOIN` won't work...
    * We need to `GROUP BY` team name and count the number of employees in each team.
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT t.name AS team_name, COUNT(e.id) AS num_employees
           FROM employees AS e RIGHT JOIN teams AS t
           ON e.team_id = t.id
           GROUP BY t.name
           ORDER BY num_employees ASC")
```

<br>

5. Do a count by year of the start dates of all employees, ordered most recent year last. <br>
[Investigate how to use the SQL `EXTRACT()` function to get the year from a date]
```{r}
dbGetQuery(db_connect, 
           "SELECT EXTRACT(YEAR FROM start_date) AS year, COUNT(id) 
           FROM employees 
           GROUP BY year 
           ORDER BY year ASC NULLS LAST")
```

<br>

6. Return the first name, last name and salary of all `employees` together with a new column called `salary_class` with value `'low'` where salary is less than 40,000 and value `'high'` where salary is greater than or equal to 40,000.<br>
[Investigate how to use the SQL `CASE()` function to return the required values `'low'` and `'high'` based on the value of salary]
```{r}
dbGetQuery(db_connect, 
           "SELECT first_name, last_name, 
           CASE 
             WHEN salary < 40000 THEN 'low' 
             ELSE 'high' 
           END AS salary_class
           FROM employees")
```

7. Only return first names where more than 1 person has the first name.

```{r}
dbGetQuery(db_connect, 
           "SELECT first_name, count(*) AS name_count
           FROM employees
           GROUP BY first_name 
           HAVING count(*) > 1
           ORDER BY name_count desc")
```

# Extensions - many-to-many problems

The following problems require joining over three tables: `employees`, `committees` and the join table `employees_committees`. Again, don't worry, take it slow and think of the join of the first two tables producing a table that we **then** join to the third table. 

<br>

1. **[Break-in problem, doesn't require any `JOIN`s]** How many of the corporation's `employees` serve in `committees`?
<details>
<summary>**Hint**</summary>
All of the details of membership of `committees` is held in a single table: `employees_committees`.
</details>
```{r}
dbGetQuery(db_connect, 
           "SELECT COUNT(DISTINCT(employee_id))
           FROM employees_committees")
```

<br>

2. Get the full details (including committee name) of any committee members based in China.
<details>
<summary>**Hints**</summary>
    * We need both joins for the many-to-many relationship with a join table.
    * We need to filter based upon the country of employees 
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT e.*, c.name AS committee_name
           FROM (employees AS e INNER JOIN employees_committees AS ec
           ON e.id = ec.employee_id)
           INNER JOIN committees AS c
           ON ec.committee_id = c.id
           WHERE e.country = 'China'")
```

<br>

3. Find the first name, last name, email address and start date of all the `employees` who are members of the 'Equality and Diversity' committee. Order the member `employees` by their length of service in the company, longest first.
```{r}
dbGetQuery(db_connect,
           "SELECT e.first_name, e.last_name, e.email, e.start_date
           FROM (employees AS e INNER JOIN employees_committees AS ec
           ON e.id = ec.employee_id)
           INNER JOIN committees AS c
           ON ec.committee_id = c.id
           WHERE c.name = 'Equality and Diversity'
           ORDER BY e.start_date ASC NULLS LAST")
```

<br>

4. **[Very tough!]** Re-use the `CASE()` operator from question 6 above to group `employees` who are members of `committees` into `salary_class` `'low'` or `'high'`. Count the number of committee members in each class.
```{r}
dbGetQuery(db_connect,
           "SELECT 
           CASE 
             WHEN salary < 40000 THEN 'low' 
             ELSE 'high' 
           END AS salary_class,
           COUNT(e.id) AS num_committee_members
           FROM (employees AS e INNER JOIN employees_committees AS ec
           ON e.id = ec.employee_id)
           INNER JOIN committees AS c
           ON ec.committee_id = c.id
           GROUP BY salary_class")
```

<br>

5. **[Very tough!]** Group committee members into the `teams` in which they work, counting the number of committee members in each team (including teams with no committee members). Order the list by the number of committee members, highest first.
<details>
<summary>**Hints**</summary>
    * This requires **three** joins: two joins for the many-to-many relationship as usual, followed by a third join to `teams`!
    * 'including teams with no committee members' - what type of join should we create to `teams`?
</details>
```{r}
dbGetQuery(db_connect,
           "SELECT t.name AS team_name, COUNT(e.id) AS num_committee_members
           FROM (employees AS e INNER JOIN employees_committees AS ec
           ON e.id = ec.employee_id)
           INNER JOIN committees AS c
           ON ec.committee_id = c.id
           RIGHT JOIN teams AS t
           ON e.team_id = t.id
           GROUP BY t.name
           ORDER BY num_committee_members DESC NULLS LAST")
```


6. How many employees are not in a committee? (This requires joining over only 2 tables)
<details>
<summary>**Hints**</summary>
    * Could you use a join and find rows without a match in the join?
</details>

```{r}
dbGetQuery(db_connect, 
           "SELECT count(*)
           FROM employees e
           LEFT JOIN employees_committees ec
           ON e.id = ec.employee_id 
           WHERE ec.employee_id IS NULL")
```


           
<hr>

```{r}
dbDisconnect(db_connect)
```
