---
title: "SQL weekend homework - Answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

<div class="blame">
author: "Del Middlemiss"<br>
date: "3rd July 2019 - rev 13th February 2020"
</div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

<hr>

<br>
<div class='emphasis'>
**Client requirements**<br>

* A client, Acme Corporation, has a PostgreSQL database containing basic details of their `employees`, the `teams` they belong to and `feedbacks` they have given on their employment (text snippets detailing positive and negative experiences, suggestions for improvement etc).
* Acme have hired us to first do some basic analysis of the data they store, and then hopefully perform a quantitative analysis of the emotional state of their corporation based on the stored `feedbacks`.

</div>
<br>

<hr>

# MVP

* Create a local PostgreSQL database called `acme_employees` and seed it using the SQL commands stored in file `acme_employees.sql`.

In `Terminal`: 

```bash
createdb acme_employees
psql -d acme_employees -f acme_employees.sql
```

* Take some time to explore your new `acme_employees` database using the `psql` command line interface (remember the `\dt` and `\d+` commands). Look for tables and the relationships between tables. How many records are there in each table? Identify any fields corresponding to primary keys and foreign keys in each table?

Three tables: `employees` (200 records), `feedbacks` (60 records) and `teams` (11 records)
(PK is 'primary key', FK is 'foreign key'):

* `employees`: PK `id`, FK `team_id` NOT NULL, referenced by `feedbacks`.
* `teams`: PK `id`, no FK, referenced by `employees`
* `feedbacks`: PK `id`, FK `employee_id` NOT NULL

Relationships:

* One team has many employees
* One employee has many feedbacks

* Create a connection from R to your local `acme_employees` database.

```{r}
library(keyring)
library(RPostgreSQL)

keyring_unlock("local")
username <- key_get(service = "postgresql_username", keyring = "local")
password <- key_get(service = "postgresql_password", keyring = "local")
keyring_lock("local")

db_connection <- dbConnect(drv = PostgreSQL(), user = username, password = password, dbname = "acme_employees", host = "localhost")

rm(username, password)
```

Now use SQL queries within `R` (i.e. use `dbGetQuery()`) to answer all of the following questions.

* Create a new `data.frame` containing the details of all `employees` held in the database

```{r}
all_employees <- dbGetQuery(db_connection,
           "SELECT * FROM employees")
all_employees
```

* Retrieve the records for all `employees` again, but this time add a field `full_name` containing the `first_name` and `last_name` of each employee separated by a space.

```{r}
dbGetQuery(db_connection,
           "SELECT *  FROM employees, 
           CONCAT(first_name, ' ', last_name) AS full_name")
```

* Retrieve the details of all `employees` again, but this time listing their team `name` (as `team_name`), rather than their `team_id`.

```{r}
dbGetQuery(db_connection,
           "SELECT e.first_name, e.last_name, e.email, t.name AS team_name
           FROM employees AS e LEFT JOIN teams AS t
           ON e.team_id = t.id")
```

* [**Bit Harder**] Do a breakdown of the number of `employees` per team (here we want two columns: `team_name` and `num_employees`). Order the table in decreasing number of employees. Include all `teams`, even if they have no employees.

```{r}
dbGetQuery(db_connection,
           "SELECT t.name AS team_name, COUNT(e.id) AS num_employees
           FROM employees AS e RIGHT JOIN teams AS t
           ON e.team_id = t.id
           GROUP BY t.name
           ORDER BY num_employees DESC")
```

* [**Bit Harder**] AcmeCorp management want to send an email survey to each of their `employees`. However, they are worried that some of the `employees` may not receive the email either because their email addresses may be invalid (in that they don't contain an '@' symbol), or they may not have a stored email address at all. Can you find all these `employees`?

```{r}
dbGetQuery(db_connection,
           "SELECT * FROM employees
           WHERE email NOT LIKE '%@%' OR email IS NULL")
```

<hr>

# Extensions

## Harder SQL queries

* Which of the `employees` provided the most recent feedback? What was the date of this feedback?

```{r}
dbGetQuery(db_connection,
           "SELECT e.id, e.first_name, e.last_name, f.date 
           FROM employees AS e INNER JOIN feedbacks AS f
           ON f.employee_id = e.id
           WHERE f.date = (SELECT MAX(date)
           FROM feedbacks)")
```

* You are chatting with one of the AcmeCorp employees, and she mentions that one of the `teams` in the corporation has some `employees` whose `first_name` contains only their first initial, e.g. "D" instead of "Dionne", "C" instead of "Charles" (they had a really formal administrator a few years back). Can you find the `name` of the team that she is most likely referring to (i.e. which team has the highest number of employees with single initial `first_names`)`?

```{r}
dbGetQuery(db_connection,
           "SELECT COUNT(*) AS num_only_first_initial, t.name AS team_name 
           FROM employees AS e LEFT JOIN teams AS t
           ON e.team_id = t.id
           WHERE e.first_name LIKE '_'
           GROUP BY team_name
           ORDER BY num_only_first_initial DESC NULLS LAST
           LIMIT 1")
```

## Combining SQL and `R` coding - basic sentiment analysis 

* AcmeCorp would like you to perform very basic 'sentiment analysis' of the stored employee feedbacks. They have provided a set of 'positive' and 'negative' words to use to score feedback messages in this analysis. 
  - The score for each message should start at zero. 
  - **Every** time a 'positive' word occurs in a message, the message's score should **increase** by 1 point.
  - **Every** time a 'negative' word occurs in a message, the message's score should **decrease** by 1 point.
  - Punctuation marks should be removed prior to scoring [**Hint** - the `gsub()` function can help here. You probably want to pass `fixed = TRUE` into this function].
  - Messages should also be converted to lowercase prior to scoring [**Hint** - the `tolower()` function can help here].
  - Write `R` code to score all of the stored feedback messages according to these rules.

```{r}
positive_words <- c("great", "good", "happy", "content", "fulfilled", "fulfilled", "positive", "positives", "cheerful", "nice", "ambition", "ambitions", "ambitious", "friendly", "progress", "joy", "fast", "achieve", "achieving", "achievements", "best", "better",  "unlimited", "pros", "competitive", "opportunity", "advancement", "friendly", "smart", "interesting", "challenging", "grow", "caring", "balance", "helpful")

negative_words <- c("worse", "worsening", "worst", "slow", "sluggish", "impatient", "unfulfilled", "unfulfilling", "negative", "bad", "limited", "limit", "limitations", "cons", "frustrated", "frustrating", "frustration", "dull", "boring", "politics", "political", "ignore", "ignored", "ignoring", "sad", "depressed", "depressing", "nasty", "stress", "stresses", "stressed", "horrid", "horrible", "terrible", "deadline", "deadlines", "uncaring", "remote")
```

```{r}
clean_message <- function(message){
  message <- tolower(message)
  # This is messy. You'll see better ways to do this later in the course
  # We can also do something like the following, using a regular expression (regex)
  # Regexes are also covered later!
  # message <- gsub(message, pattern = "[[:punct:]]", replacement = " ")
  message <- gsub(message, pattern = ",", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = ".", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = ":", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = ";", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = "!", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = "'", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = '"', replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = "(", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = ")", replacement = "", fixed = TRUE)
  message <- gsub(message, pattern = "-", replacement = " ", fixed = TRUE)
  message <- gsub(message, pattern = "/", replacement = " ", fixed = TRUE)
  message <- gsub(message, pattern = "  ", replacement = " ", fixed = TRUE)
  message <- gsub(message, pattern = "   ", replacement = " ", fixed = TRUE)
  return(message)
}
```

```{r}
score_message <- function(message, positive_words, negative_words){
  message_words <- unlist(strsplit(message, split = " ", fixed = TRUE))
  message_score <- 0
  
  for (word in positive_words){
    message_score <- message_score + sum(message_words == word)
  }
  
  for (word in negative_words){
    message_score <- message_score - sum(message_words == word)
  }
  
  return(message_score)
}
```

This is quite a tough exercise, and in reality you wouldn't do serious sentiment analysis in this way in `R`. Treat this as a coding exercise; you'll see far better approaches with dedicated text mining packages later in the course!

<br>
<details>
<summary>**Hints**</summary>

* First we obtained the `feedbacks`, with matching `employees` and their `teams` together in one dataframe.
* Next we split our solution code into two functions: `clean_message()` and `score_message()`, and ran these successively for each message stored in the database, storing the scores as we go. 
* You will want to add the scores as a column in your dataframe containing `feedbacks`, `employees` and `teams`.

Here are examples of running both functions on two messages:

```{r}
message <- "Good company, competitive salaries, nice colleagues, good location, but the food in the canteen is just terrible!"
message <- clean_message(message)
message
score_message(message, positive_words, negative_words)
```

```{r}
message <- "Really cliquey, political culture, bosses ignore any complaints (and suggestions), but good benefits package"
message <- clean_message(message)
message
score_message(message, positive_words, negative_words)
```
</details>
<br>

```{r}
# get dataframe with employees, teams and feedback messages
employees_teams_feedbacks <- dbGetQuery(db_connection,
           "SELECT e.*, f.message, t.name AS team_name
           FROM employees AS e LEFT JOIN teams AS t
           ON e.team_id = t.id INNER JOIN feedbacks AS f
           ON f.employee_id = e.id")

# go through each message in employees_teams_feedbacks, clean and score it
score <- numeric()
for (message in employees_teams_feedbacks[, 'message']){
  message <- clean_message(message)
  message_score <- score_message(message, positive_words, negative_words)
  score <- c(score, message_score)
}

# add the scores 
employees_teams_feedbacks['score'] <- score
employees_teams_feedbacks
```

* One or more `employees` may have tied in providing the most positive comment(s). Can you find `name`(s) of the `teams` these `employees` belong to? Do the same for `employees` having provided the most negative comments. Remember that we want to do this **programatically** (i.e. by coding), rather than scanning through the scores by eye.  
We can either do this in base `R`

```{r}
selector <- employees_teams_feedbacks[, 'score'] == max(employees_teams_feedbacks[, 'score']) 
mostpos_employees_teams <- employees_teams_feedbacks[selector, "team_name"]
mostpos_employees_teams

selector <- employees_teams_feedbacks[, 'score'] == min(employees_teams_feedbacks[, 'score']) 
mostneg_employees_teams <- employees_teams_feedbacks[selector, "team_name"]
mostneg_employees_teams
```

or, alternatively, we can use `sqldf`

```{r}
dbDisconnect(db_connection)
detach("package:RPostgreSQL", unload = TRUE)
library(sqldf)

teams_mostpos_employees_querystring <- "SELECT team_name
                                        FROM employees_teams_feedbacks
                                        WHERE score = (SELECT MAX(score)
                                        FROM employees_teams_feedbacks)"

mostpos_employees_teams <- sqldf(teams_mostpos_employees_querystring)
mostpos_employees_teams

teams_mostneg_employees_querystring <- "SELECT team_name
                                        FROM employees_teams_feedbacks
                                        WHERE SCORE = (SELECT MIN(score)
                                        FROM employees_teams_feedbacks)"

mostneg_employees_teams <- sqldf(teams_mostneg_employees_querystring)
mostneg_employees_teams
```




