---
title: "Reading in data homework - Answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MVP


1. Read in the `state_income_data.csv` file using the appropriate function. 

2. How do you specify your own column names when using `read_csv()`?

3. Read in the `cars.csv` file and use the `col_types()` function to ensure you load in:

* `X1` as `character`
* `mpg`, `cyl`, `disp`, `hp` as `double`
* `vs` as `logical`

4. Read in the `scotland.txt` file using the appropriate function. 

5. How do you read text into a one element character vector?

6. Read in the `steam-200k-sample.csv` file. Fix any problems you may encounter. 

7. Read in the `historic_eu_payments.tsv` file. 

8. How do you read lines of text into a multi-element text vector?

9. Read in the `africa_health_data.dta` file using the appropriate method.

10. Read in the JSON file from the following url: http://makemeanalyst.com/wp-content/uploads/2017/06/sample_json.txt

<details>
<summary>**MVP Answers**</summary>

```{r, eval = FALSE}
# 1. Read in the `state_income_data.csv` file using the appropriate function. 
income <- read_csv("data/state_income_data.csv")

# 2. How do you specify your own column names when using `read_csv`?
col_names =

# 3. Read in the `cars.csv` file and use the 'col_types' function to ensure you load in:
# `X1` as character, `mpg`, `cyl`, `disp`, `hp` as double, `vs` as logical
car_data <- read_csv("data/cars.csv", col_types = cols( X1 = col_character(), 
                     mpg = col_double(),
                     cyl = col_double(),
                     disp = col_double(),
                     hp = col_double(),
                     vs = col_logical()))

# 4. Read in the `scotland.txt` file using the appropriate function.
## need to use delim one as it's a delimited file 
scotland <- read_delim("data/scotland.txt", delim = ";")

# 5. How do you read text into a one element character vector?
read_file

# 6. Read in the `steam-200k-sample.csv` file. Fix any problems you may encounter. 
# load in the video games data, and remove the metadata 
video_games <- read_csv("data/steam-200k-sample.csv",skip = 4)

# 7. Read in the `historic_eu_payments.tsv` file. 
eu_payments <- read_tsv("data/historic_eu_payments.tsv")

# 8. How do you read lines of text into a multi-element text vector?
read_lines

# 9. Read in the `africa_health_data.dta` file using the appropriate method.
library(haven)
africa_health <- read_stata("data/africa_health_data.dta")
head(africa_health)

# 10. Read in the json file from the following url: http://makemeanalyst.com/wp-content/uploads/2017/06/sample_json.txt

library("rjson")
sample_json_data <- fromJSON(file="http://makemeanalyst.com/wp-content/uploads/2017/06/sample_json.txt")
```
</details>

# Extension

A lot of Excel sheets you will encounter will contain more than one sheet. Try to write some code that reads in all the different sheets in the file `green_bindle_consulting.xlsx`. 

[**Hint**: you will need to write a `for` loop]

<details>
<summary>**Extension Answers**</summary>

```{r, eval = FALSE}

library(readxl)

## Method 1 - loops & naming the sheets 'data_sheet_1' etc.
num_sheets<-length(excel_sheets("data/green_bindle_consulting.xlsx"))

for(i in 1:num_sheets) {
  assign(paste0("data_sheet_",i), data.frame(read_excel("data/green_bindle_consulting.xlsx", 
                                                sheet = i)))
}
## Method 2 - loops & naming the sheets the names of the tabs.
name_sheets<-excel_sheets("data/green_bindle_consulting.xlsx")
num_sheets<-length(name_sheets)

for(i in 1:num_sheets) {
  assign(name_sheets[i], data.frame(read_excel("data/green_bindle_consulting.xlsx", 
                                                        sheet = i)))
  }


## Method 3 - lappy & naming the sheets the names of the tabs.
name_sheets<-excel_sheets("data/green_bindle_consulting.xlsx")
num_sheets<-length(name_sheets)

all_tabs <- lapply(name_sheets, function(x) read_excel("data/green_bindle_consulting.xlsx", sheet = x))
names(all_tabs)<-name_sheets
#list2env(all_tabs, globalenv()) #may not be necessary to do this step - may want to just call each element of the list

## Method 4 - start with a list and loop through assigning elements of the list
all_tabs <- list()
name_sheets<-excel_sheets("data/green_bindle_consulting.xlsx")

for(sheet in name_sheets) {
  all_tabs[[sheet]] <- data.frame(read_excel("data/green_bindle_consulting.xlsx", 
                                                    sheet = sheet))
}


