---
title: "Shiny Lab - Answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

<div class="blame">
author: "Mhairi McNeill"<br>
date: "24/06/2019"
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

**Learning Objectives**<br>

* Get practice applying everything you've learnt about Shiny apps
* Get used to creating an app from a design.

**Duration - 3 hours**<br>

In this lab you are going to be shown several apps, and we want you to recreate them as closely as you can. Throughout this lab you'll use the `students_big` dataset from the Code Clan Data package. You can think of the image you see as the first stage in our workflow - creating a diagram. Go though the other stages and build up the app.

# Height and Arm Span vs Age

In this app we are plotting two histograms that show the height and arm-span for children of different ages, as chosen by radio buttons on the top of the app.

Hints:

- Use `radioButtons` with `inline = TRUE`, to get horizontal radio buttons.
- This app uses `fluidRow` and `column`
- Both plots use the same data. Can you use `reactive` to make this efficient?

```{r, eval = FALSE}
library(CodeClanData)
library(ggplot2)
library(dplyr)
library(shiny)

ui <- fluidPage(
  radioButtons("age", "Age", choices = sort(unique(students_big$ageyears)), inline = TRUE),
  fluidRow(
    column(6,
      plotOutput("height")
    ),
    column(6,
      plotOutput("arm_span")
    )
  )
)

server <- function(input, output) {
  
  filtered_df <- reactive({
    students_big %>%
      filter(ageyears == input$age)
  })
  
  output$height <- renderPlot({
    ggplot(filtered_df()) +
      aes(x = height) +
      geom_histogram()
  })
  
  output$arm_span <- renderPlot({
    ggplot(filtered_df()) +
      aes(x = arm_span) +
      geom_histogram()
  })


}

shinyApp(ui, server)
```

# A customisable scatter plot

Here we are going to change attributes of a scatter plot, without changing the data used to make the scatter plot.

Hints:

- Use a named list inside an input, to display different values that the input takes. For example, for our colour input you will want `choices = c(Blue = "#3891A6", Yellow = "#FDE74C", Red = "#E3655B")`
- The choices for shape are: Square = 15, Circle = 16, Triangle = 17. You will need to force the the input value to be numeric inside the graph. Outputs of `selectInput` are character by default. 
- `ggtitle` is the easiest way to add the title to 
- This uses the sidebar layout


```{r, eval = FALSE}
library(CodeClanData)
library(ggplot2)
library(dplyr)
library(shiny)

ui <- fluidPage(
  
  titlePanel("Reaction Time vs. Memory Game"),
  
  sidebarLayout(
    sidebarPanel(
      radioButtons("colour",
        "Colour of points",
        choices = c(Blue = "#3891A6", Yellow = "#FDE74C", Red = "#E3655B")
      ),

      sliderInput("alpha",
        "Transparency of points",
        value = 0.7,  max = 1, min = 0
      ),
      
      selectInput("shape",
        "Shape of Points",
        choices = c(Square = 15, Circle = 16, Triangle = 17)
      ),
      
      textInput("title",
        "Title of Graph",
        value = "Reaction Time vs. Memory Game"
      )
    ),

    mainPanel(
      plotOutput("scatter_plot")
    )
  )
)


server <- function(input, output) {
  
  output$scatter_plot <- renderPlot({
    
    ggplot(students_big) +
      aes(x = reaction_time, y = score_in_memory_game) +
      geom_point(
        colour = input$colour,
        alpha  = input$alpha,
        shape  = as.numeric(input$shape)
      ) +
      labs(title = input$title)
    
  })

}

shinyApp(ui, server)
```

# Three different plot options

The app has three different options to show the same data: a bar chart, a pie chart and a stacked bar chart.

Hints:

- You will need to use if-statements inside server to do this. You can use if statements inside `renderPlot`. Remember that the last thing `renderPlot` should return is the plot you want to make.

```{r, eval = FALSE}
library(CodeClanData)
library(ggplot2)
library(dplyr)
library(shiny)

ui <- fluidPage(
  
  radioButtons("type",
    "Plot Type",
    choices = c("Bar", "Pie Chart", "Stacked Bar")
  ),
  
  plotOutput("plot")
   
)

server <- function(input, output) {
  
  output$plot <- renderPlot({
    
    if (input$type == "Bar"){
      
      plot <- 
      ggplot(students_big) +
        aes(x = gender, fill = gender) +
        geom_bar()
      
    }
    
    if (input$type == "Pie Chart"){
      
      plot <- 
      ggplot(students_big) +
        aes(x = 1, fill = gender) +
        geom_bar() +
        coord_polar("y")
      
    }
    
    if (input$type == "Stacked Bar"){
      
      plot <-
      ggplot(students_big) +
        aes(x = 1, fill = gender) +
        geom_bar() 
      
    }
    
    plot
  })
  
}

shinyApp(ui, server)
```

# An app with tabs

This Shiny application has tabs, here's how it looks in both tabs:

Hints:

- There's lots of different regions in this data set. You will probably want to use `unique` to find the different regions in the data to use in the input.
- The plots and tables don't update until the button "Update Plots and Table" is pressed.
- The plot and the data both use the same dataset.
- The table in the data table uses `dataTableOutput`.
- The two plots here are done with two separate plot outputs. However, it would be possible to create this using faceting.
- This app uses sidebar layout, tab panels and fluid row (for the plots).


```{r, eval = FALSE}
library(CodeClanData)
library(ggplot2)
library(dplyr)
library(shiny)

ui <- fluidPage(
  
  titlePanel("Comparing Importance of Internet Access vs. Reducing Pollution"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("gender",
                  "Gender",
                  choices = c(Male = "M", Female = "F")
      ),
      
      selectInput("region",
                  "Region",
                  choices = unique(students_big$region)
      ),
      
      actionButton("go", "Generate Plots and Table")
      
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("Plots",
                 fluidRow(
                   column(6, plotOutput("internet_access")),
                   column(6, plotOutput("reducing_pollution"))
                 )
    
        ), 
        
        tabPanel("Data",
                dataTableOutput("table") 
        )
        
      )
    )
  )
)
  

server <- function(input, output) {
  
  filtered_data <- eventReactive(input$go, {
    students_big %>%
      filter(gender == input$gender) %>%
      filter(region == input$region) %>%
      select(
        region,
        gender,
        importance_internet_access,
        importance_reducing_pollution
      )
  })
  
  output$internet_access <- renderPlot({
    ggplot(filtered_data()) +
      aes(x = importance_internet_access) +
      geom_histogram()
  })
  
  output$reducing_pollution <- renderPlot({
    ggplot(filtered_data()) +
      aes(x = importance_reducing_pollution) +
      geom_histogram()
  })
  
  output$table <- renderDataTable({
    filtered_data()
  })
  
}

shinyApp(ui, server)
```

# A trickier one - how do we change `ggplot2` variables?

Here we have a basic app with only one plot. What's tricky is that we can choose between plotting each of the 6 "importance" variables.

Hints:

- When you select a variable using `selectInput`, it returns a string instead of a name.
- If you use `aes_string` instead of `aes`, then you can accept strings instead of name in `ggplot`.
- This plots are made with `geom_density`, with `alpha = 0.5`.
- For more complicated situations like this - when you are working with names, rather than values - you might need to learn about `tidyeval`. 
- Here's some resources on `tidyeval` if you are interested:
  * short video https://www.youtube.com/watch?v=nERXS3ssntw
  * section of book https://tidyeval.tidyverse.org/sec-why-how.html 


```{r, eval = FALSE}
library(CodeClanData)
library(ggplot2)
library(dplyr)
library(shiny)

ui <- fluidPage(
  
  selectInput("variable",
              "Which variable to plot?",
              choices = c(
                "importance_reducing_pollution",
                "importance_recycling_rubbish",
                "importance_conserving_water", 
                "importance_saving_enery",
                "importance_owning_computer",
                "importance_internet_access"
              )
  ),
  plotOutput("plot")
)
  

server <- function(input, output) {

  
  output$plot <- renderPlot({
    
    ggplot(students_big) +
      aes_string(x = input$variable, fill = "gender") +
      geom_density(alpha = 0.5)
    
  })
  
}

shinyApp(ui, server)
```


