---
title: "Probability quiz"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../resources/note_styles.css
  pdf_document: default
---

<div class="blame">
author: "Del Middlemiss"<br>
date: "26th June 2019"
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

Today was pretty full-on! Here are some questions to help you consolidate what you learned. Feel free to work on paper or in R, whichever you are most comfortable with:

```{r, message = FALSE}
library(prob)
library(gtools)
library(tidyverse)
```

<br>
1. If I roll a six-sided dice once, what is the probability of getting a number less than 5?

<br>
2. How many outcomes are there in the sample space of an experiment in which we toss three dice?

<br>
3. In an experiment where we roll four dice:

  i. how many outcomes are in the event `total_equals_5_6_7` (i.e. the total of the four dice equals $5$, $6$ or $7$)?
  ii. what is the probability of this event?
  iii. what assumptions do we apply in answering part iii?

<details>
<summary>**Hint**</summary>
For part i, `rowSums(.)` may help you setup up the necessary random variable
</details>

<br>
4. Five chairs are placed in a row in a room, along with five people of differing height. The people sit down randomly on the chairs. What is the probability that they sit in order of decreasing height, left to right?

<details>
<summary>**Hint**</summary>
This is an arrangement where order matters. Of all the possible arrangements, how many are correct?
</details>

<br>
5. I roll six dice:

  i. what is the probability I see the numbers 6, 5, 4, 3, 2 and 1 in order on the dice, left to right?
  ii. what is the probability I see one each of the numbers 6, 5, 4, 3, 2 and 1 on the dice regardless of order?
  
[**Hint** have a look at the `isin()` function in the `prob` package. For example, to `filter` all rows containing $6$, $1$, $2$ in that order, you could use `isin(., c(6, 1, 2), ordered = TRUE)`]
  
<br>
6. If I draw a single card from a 52-card deck, and it is black, what is the probability that its suit is spades? <br>

<details> 
<summary>**Hints**</summary>

* Investigate the `cards()` function in the `prob` package, if you'd like to try this in R (make a deck without jokers). 
* You could think about adding your own `black_card` random variable to the sample space, which would be `TRUE` if the `suit` is 'Spade' or 'Club', and `FALSE` otherwise.

</details>

<br>
7. Eight patients: Alice, Bob, Carmen, Dinesh, Ella, Fergus, Gianna and Henry are waiting to see a doctor, but only five appointment slots are left today. 

  i. In how many different orders can the doctor see patients today?<br>
  <details> 
  <summary>**Hint**</summary>
  `permutations()` returns a matrix, but it's better to convert it to a tibble with `as_tibble()`. You can then set column names etc.]
  </details>
  
  ii. What is the probability of Gianna and Danesh **both** being seen today?<br>
  <details> 
  <summary>**Hint**</summary> 
  Filter using `isin()`
  </details>
  
  iii. What is the probability of Gianna being seen **before** Dinesh today?<br>
    <details> 
  <summary>**Hint**</summary>
  Think about the `ordered =` argument of `isin()`
  </details>
  
  iv. **(harder)** Carmen and Henry have now been triaged as high priority and will be seen in the first two available slots, although their order hasn't yet been determined. In how many different orders can the patients now be seen?<br>
    <details> 
  <summary>**Hint**</summary>
  Think of this as a two-step experiment: (1) choose order of high priority patients, (2) choose order of remaining patients. Get separate tibbles for steps (1) and (2) and then `merge()` them (make sure the column names are all distinct however or `merge()` will try to join on any repeated column names - not what we want).]
  </details>
  

<br>
8. If order is not significant, how many ways are there to select a 5 student quiz team from a group of 8 students?

<br>
9. **(harder)** How would your answer for question 8 change if a team captain had to be selected for each team?


<br>
10. The following data from a sample of 120 families shows the record of university attendance by mothers and their eldest sons.

  - in 32 families, both the mother and son attended university; 
  - in 31 families, neither the mother nor the son attended university
  - in 22 families, the mother attended university while the son did not
  - in 35 families, the son attended university while the mother did not.

Arrange the data above in a convenient dataframe and answer the following questions:

  i. What is the probability that a son attended university given that his mother attended university?
  ii. What is the probability that a son attended university given that his mother did not attend university?
  iii. Is attending university by the son *independent* of whether his mother attended university?

<details>
<summary>**Hint - code to set up dataframe - try yourself first though!**</summary>
```{r, eval = FALSE}
uni <- data.frame(
  mother_yes = c(32, 22), 
  mother_no = c(35, 31)
)
uni <- cbind(uni, rowSums(uni))
uni <- rbind(uni, colSums(uni))
colnames(uni) <- c("mother_yes", "mother_no", "sum")
rownames(uni) <- c("son_yes", "son_no", "sum")
```
</details>

