---
title: "Probability quiz - solutions"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../resources/note_styles.css
  pdf_document: default
---

<div class="blame">
author: "Del Middlemiss"<br>
date: "26th June 2019"
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

Today was pretty full-on! Here are some questions to help you consolidate what you learned. Feel free to work on paper or in R, whichever you are most comfortable with:

```{r, message = FALSE}
library(prob)
library(gtools)
library(tidyverse)
```

<br>
1. If I roll a six-sided dice once, what is the probability of getting a number less than 5?

```{r}
S <- rolldie(1)
S_less_5 <- S %>%
  filter(X1 < 5)

prob <- nrow(S_less_5) / nrow(S)
prob
```

<br>
2. How many outcomes are there in the sample space of an experiment in which we toss three dice?

```{r}
# can either do this using the multiple step counting rule
size_S <- 6 * 6 * 6
# or just brute force using one of the `prob` functions
size_S <- nrow(rolldie(3))
size_S
```

<br>
3. In an experiment where we roll four dice:

  i. how many outcomes are in the event `total_equals_5_6_7` (i.e. the total of the four dice equals $5$, $6$ or $7$)?
  ii. what is the probability of this event?
  iii. what assumptions do we apply in answering part iii?

<details>
<summary>**Hint**</summary>
For part i, `rowSums(.)` may help you setup up the necessary random variable
</details>

```{r}
# i.
S <- rolldie(4)

# add total random variable
S <- S %>%
  mutate(total = rowSums(.))

S_total_between_5_7 <- S %>%
  filter(total %in% c(5, 6, 7))

nrow(S_total_between_5_7)

# ii.
prob <-nrow(S_total_between_5_7) / nrow(S)
prob

# iii.
# We assume all outcomes in the sample space are equally likely!
# i.e. the dice are all unbiased
```

<br>
4. Five chairs are placed in a row in a room, along with five people of differing height. The people sit down randomly on the chairs. What is the probability that they sit in order of decreasing height, left to right?

<details>
<summary>**Hint**</summary>
This is an arrangement where order matters. Of all the possible arrangements, how many are correct?
</details>

```{r}
# number of ordered arrangements:
num_outcomes <- factorial(5)
# only one of these is correctly ordered:
prob_perfect_order <- 1/num_outcomes
prob_perfect_order
```

<br>
5. I roll six dice:

  i. what is the probability I see the numbers 6, 5, 4, 3, 2 and 1 in order on the dice, left to right?
  ii. what is the probability I see one each of the numbers 6, 5, 4, 3, 2 and 1 on the dice regardless of order?
  
[**Hint** have a look at the `isin()` function in the `prob` package. For example, to `filter` all rows containing $6$, $1$, $2$ in that order, you could use `isin(., c(6, 1, 2), ordered = TRUE)`]
  
```{r}
S <- rolldie(6)

# i. only one outcome has 6, 5, 4, 3, 2, 1 in order so prob
prob <- 1 / nrow(S)
prob

# or
S_in_order <- S %>%
  filter(isin(., c(6, 5, 4, 3, 2, 1), ordered = TRUE))
prob <- nrow(S_in_order) / nrow(S)
prob

# or 'mathematically': there are 6^6 outcomes on roll of six dice, of which only one is correct
prob <- 1 / 6^6
prob

# ii.
# row must contain 6, 5,..., 1, but can occur in any order
S_contains_654321 <- S %>%
  filter(isin(., c(6, 5, 4, 3, 2, 1), ordered = FALSE))
prob <- nrow(S_contains_654321) / nrow(S)
prob

# 'mathematically': each outcome has 1/6^6 probability, but now 6, 5,..., 1 can occur in any order without replacement
# there are 6! orders of six objects without replacement
# So prob equals 6! times the probability of a single outcome, i.e. 6! / 6^6
prob <- factorial(6) / 6^6
prob
```

<br>
6. If I draw a single card from a 52-card deck, and it is black, what is the probability that its suit is spades? <br>
<details> 
<summary>**Hints**</summary>

* Investigate the `cards()` function in the `prob` package, if you'd like to try this in R (make a deck without jokers). 
* You could think about adding your own `black_card` random variable to the sample space, which would be `TRUE` if the `suit` is 'Spade' or 'Club', and `FALSE` otherwise.

</details>

```{r}
# this is a conditional probability problem.
# possible source of confusion here: the conditioning event comes first
S <- cards(jokers = FALSE)

S <- S %>%
  mutate(black_card = suit == "Club" | suit == "Spade")

S_spades_and_black <- S %>%
  filter(black_card == TRUE & suit == "Spade")
prob_spades_and_black <- nrow(S_spades_and_black) / nrow(S)

S_black <- S %>%
  filter(black_card == TRUE)
prob_black <- nrow(S_black) / nrow(S)

prob_spades_given_black <- prob_spades_and_black / prob_black
prob_spades_given_black

# or 'mathematically', i.e. intuitively in this case
# 'it is black' limits event space to 52/2 = 26 cards
# next, only two suits are black, so card will be a spade 13/26 of the time, i.e. 1/2
# in plain language, IF we know the card is black, it will be a spade half the time
```

<br>
7. Eight patients: Alice, Bob, Carmen, Dinesh, Ella, Fergus, Gianna and Henry are waiting to see a doctor, but only five appointment slots are left today. 

  i. In how many different orders can the doctor see patients today?<br>
  <details> 
  <summary>**Hint**</summary>
  `permutations()` returns a matrix, but it's better to convert it to a tibble with `as_tibble()`. You can then set column names etc.]
  </details>
  
  ii. What is the probability of Gianna and Danesh **both** being seen today?<br>
  <details> 
  <summary>**Hint**</summary> 
  Filter using `isin()`
  </details>
  
  iii. What is the probability of Gianna being seen **before** Dinesh today?<br>
  <details> 
  <summary>**Hint**</summary>
  Think about the `ordered =` argument of `isin()`
  </details>
  
  iv. **(harder)** Carmen and Henry have now been triaged as high priority and will be seen in the first two available slots, although their order hasn't yet been determined. In how many different orders can the patients now be seen?<br>
  <details> 
  <summary>**Hint**</summary>
  Think of this as a two-step experiment: (1) choose order of high priority patients, (2) choose order of remaining patients. Get separate tibbles for steps (1) and (2) and then `merge()` them (make sure the column names are all distinct however or `merge()` will try to join on any repeated column names - not what we want).]
  </details>
  
```{r}
# i.
# this is an 8P5 problem: permute 5 patients from 8.
patients <- c("Alice", "Bob", "Carmen", "Dinesh", "Ella", "Fergus", "Gianna", "Henry")
S <- as_tibble(permutations(n = 8, r = 5, v = patients, repeats.allowed = FALSE))
names(S) <- c("appt1", "appt2", "appt3", "appt4", "appt5")
# answer
nrow(S)

# ii.
# now filter the tibble
S_Gianna_Dinesh <- S %>%
  filter(isin(., c("Gianna", "Dinesh"), ordered = FALSE))
# then calculate required probability
prob <- nrow(S_Gianna_Dinesh) / nrow(S)
prob

# iii.
# filter tibble, this time with ordering, Gianna before Dinesh
S_Gianna_before_Dinesh <- S %>%
  filter(isin(., c("Gianna", "Dinesh"), ordered = TRUE))
# new prob
prob <- nrow(S_Gianna_before_Dinesh) / nrow(S)
prob
# this should be half the probability of question ii, because, in filtered outcomes for 
# question ii., Gianna would be before Dinesh half the time.

# iv.
# first step is a 2P2 problem: permute 2 patients from 2.
high_priority_patients <- c("Carmen", "Henry")
S_high_priority <- as_tibble(permutations(n = 2, r = 2, v = high_priority_patients, repeats.allowed = FALSE))
names(S_high_priority) <- c("appt1", "appt2")

# second step is a 6P3 problem: permute 3 patients from 6.
normal_priority_patients <- c("Alice", "Bob", "Dinesh", "Ella", "Fergus", "Gianna")
S_normal_priority <- as_tibble(permutations(n = 6, r = 3, v = normal_priority_patients, repeats.allowed = FALSE))
names(S_normal_priority) <-c("appt3", "appt4", "appt5")

# merge these dataframes
S <- merge(S_high_priority, S_normal_priority)
# final answer
nrow(S)

# How would we do this 'mathematically': high priority patients: 2! = 2 ways to order them. Now have 6 remaining patients for 3 named slots, so have 6!/3! ways to order. Use multiple step counting rule: (6 * 5 * 4) * (2 * 1) = 240 
```

<br>
8. If order is not significant, how many ways are there to select a 5 student quiz team from a group of 8 students?
```{r}
teams <- combinations(n = 8 , r = 5, v = 1:8, repeats.allowed = FALSE)
nrow(teams)
```

<br>
9. **(harder)** How would your answer for question 8 change if a team captain had to be selected for each team?

```{r}
# step 1: choose a captain. We have 8 choices
num_captains <- 8

# step 2: for each choice of captain, choose 4 remaining team members from 7 
teams <- combinations(n = 7, r = 4, v = 1:7, repeats.allowed = FALSE)
num_team_combinations <- nrow(teams)

# combinations
num_captains * num_team_combinations

# 'mathematically': choice of 8 captains, and then, for each captain, a choice of 4 team members
# from 7 students, i.e. 7C4 = 7!/(4!*3!) = 7*6*5/(3*2*1) = 35. 
# So final number of team variations with captain is 8 * 35 = 280

# alternative 'mathematical' method. In question 8, worked out that there are 56 ways to choose a team of 5 from amongst 8 students. Now, for each of those 56 teams, there are 5 choices of captain, so final number of team variations with captain is 56 * 5 = 280
```

<br>
10. The following data from a sample of 120 families shows the record of university attendance by mothers and their eldest sons.

  - in 32 families, both the mother and son attended university; 
  - in 31 families, neither the mother nor the son attended university
  - in 22 families, the mother attended university while the son did not
  - in 35 families, the son attended university while the mother did not.

Arrange the data above in a convenient dataframe and answer the following questions:

  i. What is the probability that a son attended university given that his mother attended university?
  ii. What is the probability that a son attended university given that his mother did not attend university?
  iii. Is attending university by the son *independent* of whether his mother attended university?

<details>
<summary>**Hint - code to set up dataframe - try yourself first though!**</summary>
```{r, eval = FALSE}
uni <- data.frame(
  mother_yes = c(32, 22), 
  mother_no = c(35, 31)
)
uni <- cbind(uni, rowSums(uni))
uni <- rbind(uni, colSums(uni))
colnames(uni) <- c("mother_yes", "mother_no", "sum")
rownames(uni) <- c("son_yes", "son_no", "sum")
```
</details>

```{r}
uni <- data.frame(
  mother_yes = c(32, 22), 
  mother_no = c(35, 31)
)
uni <- cbind(uni, rowSums(uni))
uni <- rbind(uni, colSums(uni))
colnames(uni) <- c("mother_yes", "mother_no", "sum")
rownames(uni) <- c("son_yes", "son_no", "sum")

# total number of families
total <- uni["sum", "sum"]
total

# marginal probabilities
prob_son_yes <- uni["son_yes", "sum"] / total
cat("prob_son_yes: ", prob_son_yes)
prob_son_no <- uni["son_no", "sum"] / total
cat("prob_son_no: ", prob_son_no)
prob_mother_yes <- uni["sum", "mother_yes"] / total
cat("prob_mother_yes: ", prob_mother_yes)
prob_mother_no <- uni["sum", "mother_no"] / total
cat("prob_mother_no", prob_mother_no)

# i.
prob_son_and_mother_yes <- uni["son_yes", "mother_yes"] / total
prob_son_yes_given_mother_yes <- prob_son_and_mother_yes / prob_mother_yes
cat("prob_son_yes_given_mother_yes: ", prob_son_yes_given_mother_yes)

# ii.
prob_son_yes_and_mother_no <- uni["son_yes", "mother_no"] / total
prob_son_yes_given_mother_no <- prob_son_yes_and_mother_no / prob_mother_no
cat("prob_son_yes_given_mother_no: ", prob_son_yes_given_mother_no)

# iii.
# The probability of a son going to university depends upon whether their mother
# went to university, as 
prob_son_yes_given_mother_yes != prob_son_yes
# and
prob_son_yes_given_mother_no != prob_son_yes
```





