---
title: "Connections homework - answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

<div class="blame">
author: "Del Middlemiss"<br>
date: "12th June 2019"
</div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# MVP

## Local PostgreSQL connections

* Create a local PostgreSQL database called `wrestling`
```bash
createdb wrestling
```

<br>

* Seed the `wrestling` database from the file `wrestling.sql` using the `psql` PostgreSQL command line interface.
```bash
psql -d wrestling -f wrestling.sql
```

<br>

* Create a connection to the local `wrestling` database from R. 
    - Don't forget to lock all the keychains you use.
    - Pass the argument `max.con = 2` to `PostgreSQL()` to make sure you don't have too many simultaneous connections. We'll be setting up a local connection and a remote connection, hence '2'.
    - Remember to add a separate 'disconnect' block of code. We'll write queries between the 'connect' and 'disconnect' blocks.
```{r}
library(keyring)
library(RPostgreSQL)

keyring_unlock("local")
username <- key_get("postgresql_username", keyring = "local")
password <- key_get("postgresql_password", keyring = "local")
keyring_lock("local")

db_connect <- dbConnect(PostgreSQL(max.con = 2), user = username, password = password, dbname='wrestling', host='localhost')

rm(username, password)
```

```{r, eval=FALSE, echo=TRUE}
dbDisconnect(db_connect)
```

<br>

* Using the `R` connection you just created, write and execute a SQL query to return all the wrestlers in the `wrestlers` table.
```{r}
dbGetQuery(db_connect, "SELECT * FROM wrestlers")
```

<br>

* Explore the columns stored in the table. State which data type you think each column stores.

`id` and `total_matches` are `INT`<br>
`born`, `name` and `gender` are `VARCHAR`<br>
`win_percent` is `REAL` or `FLOAT`, a continuous numerical type.

<br>

## Remote PostgreSQL connections

Now we'll try to connect to a remote PostgreSQL database holding the same data. As earlier, this connection may or may not work, depending on how many other connections to the server are currently active. But try a few times, and don't worry if it proves difficult. Being able to write the connection code is what we're looking for here!

Here are the details of the remote PostgreSQL database we wish to connect to:

    username: 'mvhbdgdk'
    password: 'Ejn5vEqOUfCA1Cxg1rqkh9NY5VbKKAIm'
    database_name: 'mvhbdgdk'
    host: 'manny.db.elephantsql.com' or '34.245.37.126' (these are equivalent)

Do the following:

* Store the credentials above in appropriately named keys in the `remote` keyring.

Do the following in the console in RStudio, or use `Keychain Access`
```{r, eval=FALSE, echo=TRUE}
# in the R console in RStudio
keyring_unlock("remote")
key_set("wrestling_postgresql_username", keyring = "remote")
# enter 'mvhbdgdk' at the prompt
key_set("wrestling_postgresql_password", keyring = "remote")
# enter 'Ejn5vEqOUfCA1Cxg1rqkh9NY5VbKKAIm' at the prompt
keyring_lock("remote")
```

<br>

* Use your credentials to establish a connection from `R` to the remote database.
```{r}
keyring_unlock("remote")
username <- key_get("wrestling_postgresql_username", keyring = "remote")
password <- key_get("wrestling_postgresql_password", keyring = "remote")
keyring_lock("remote")

db_connect_remote <- dbConnect(drv = PostgreSQL(max.con = 2), user = username, password = password, dbname='mvhbdgdk', host='34.245.37.126')

rm(username, password)
```

<br>

* Remember to lock keychains and write appropriate code to close database connections, as before.
```{r, eval=FALSE, echo=TRUE}
dbDisconnect(db_connect_remote)
```

<br>

* Using this `R` connection, write and execute a query to return all the wrestlers from the `wrestlers` table
```{r}
dbGetQuery(db_connect_remote, "SELECT * FROM wrestlers")
```

<hr>

# Extensions

Return to using your `R` connection to the local `wrestlers` database. Write and run queries to solve the following more difficult problems:

<br>

* Return all the details for any wrestlers from the United Kingdom
```{r}
dbGetQuery(db_connect, 
           "SELECT *
           FROM wrestlers
           WHERE born = 'United Kingdom'")
```

<br>

* Get the names of all Japanese wrestlers who have appeared in 40 or more matches.
```{r}
dbGetQuery(db_connect, 
           "SELECT *
           FROM wrestlers
           WHERE born = 'Japan' AND total_matches >= 40")
```

<br>

* Find the wrestlers who have won between 50% and 60% of their matches.
```{r}
dbGetQuery(db_connect, 
           "SELECT *
           FROM wrestlers
           WHERE win_percent BETWEEN 50 AND 60")
```

<br>

* Find all the wrestlers with names beginning with 'X'.
```{r}
dbGetQuery(db_connect, 
           "SELECT *
           FROM wrestlers
           WHERE name LIKE 'X%'")
```

<br>

* Find all the wrestlers who were born in countries beginning with the letter 'I'. Try this using `IN` (there are 9 countries) and using a wildcard comparison. 
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM wrestlers 
           WHERE born IN ('Ireland', 'Italy', 'Iran', 'Iraq', 'Iceland', 'Israel', 'India', 'Indonesia', 'Ivory Coast')")

dbGetQuery(db_connect, 
           "SELECT * 
           FROM wrestlers 
           WHERE born LIKE 'I%'")
```

<br>

* Find all the wrestlers who were **not** born in countries beginning with the letter 'I'.
```{r}
dbGetQuery(db_connect, 
           "SELECT * 
           FROM wrestlers 
           WHERE born NOT LIKE 'I%'")
```

```{r}
dbDisconnect(db_connect)
dbDisconnect(db_connect_remote)
```






